---
title: "PHOBILE frequently asked questions"
layout: default
---

This page describes some of the basic properties of the PHOIBLE database,
including XXX, YYY, ZZZ.

```{r opts, echo=FALSE}
knitr::opts_chunk$set(fig.path = "images/faq/")
```

# How we classify inventories

For the most part, every phonological inventory in PHOIBLE is based on one and
only one language description (usually a research article, book chapter,
dissertation, or descriptive grammar). The conceptual term for this is
"doculect" — an instance of documentation of an instance of linguistic behavior
at a particular time and place. Occasionally, a single document may provide
information about multiple doculects; for example, a dissertation that described
and compared 3 related dialects or languages spoken in a particular region. In
that case, 3 phonological inventories in PHOIBLE corresponding to 3 doculects
might all be linked to the same document, but each inventory is still linked to
only one document (it just happens to be *the same document* for those three
inventories).

An exception to the "one document per inventory" rule arises for inventories
that were originally part of a curated phonological database such as UPSID or
SPA. In those collections, inventories were often based on multiple descriptions
of linguistic behavior, written by different linguists; those descriptions were
believed to be describing the same language, and disagreements between the
descriptions were adjudicated by the experts who compiled the collection.

Since most PHOIBLE inventories (aside from some UPSID or SPA ones, as mentioned
above) are based on a single document, it is fairly straightforward to link each
PHOIBLE inventory to Glottolog, which provides links between linguistic
description documents and unique identifiers for dialects, languages, and
groupings of dialects and languages at various levels (the preferred term for
any of these levels of specificity / grouping is "languoid"). Thus in PHOIBLE it
is usually the case that a single inventory corresponds to a single languoid,
and in most cases that languoid is a "leaf node" in the Glottolog tree, i.e., it
represents a particular dialect known to be used at a particular place and time
(rather than a group of dialects, a language, or a language family). However, in
the few cases where multiple document sources were consulted for an inventory,
it may not be possible to link that inventory to a unique Glottolog leaf node.
In such cases, the inventory in PHOIBLE is linked to the lowest possible
Glottolog node that dominates the leaf nodes of each source document. For
example, XXX TODO GIVE A REAL-LIFE EXAMPLE HERE XXX.

In addition to providing Glottolog languoid codes ("glottocodes") for each
inventory, PHOIBLE also includes ISO 639-3 language codes ("isocodes") for each
inventory. The link between glottocodes and isocodes is provided by Glottolog
(PHOIBLE no longer maintains a separate mapping).

# Duplicate entries in PHOIBLE

In light of the above discussion of how we classify phonological inventories, it
should come as no surprise that PHOIBLE can contain multiple inventories for
"the same" languoid, depending on what kind of languoid you're interested in.
All it takes is the existence of two or more descriptive documents associated
with the same group of speakers, where "group of speakers" is ultimately
determined by your desired level of granularity. To give a concrete example, one
researcher may be interested in comparing "languages", and so might wish to
treat all inventories of "English" as duplicates in some sense, regardless of
the regional dialect, sociolect, etc. represented in the doculect and encoded in
the phonological inventory. Such a researcher might *filter* or *sample* the
PHOIBLE dataset in order to include only one inventory for each isocode (how she
might choose to implement that filter is a separate question; see "Filtering and
sampling inventories" for examples). Other researchers may not care about
"duplicates" in that sense, and may choose to include all inventories in their
analysis (or, they may filter the dataset to include only inventories with a
particular feature of interest such as breathy-voiced vowels).

Because a key principle of PHOIBLE is to remain as faithful as possible to the
description of speech sounds in the original doculects (though not necessarily
to the exact *symbols* used in the document, if they don't reflect the textual
descriptions), it may arise that linguists disagree on their phonological
analysis of a particular languoid, and hence there may be two PHOIBLE
inventories (with unique inventory IDs) that link to the same glottocode (but
different documents) but nonetheless contain different sets of phonemes.

TODO: CONCRETE EXAMPLE OF TWO INVENTORIES WITH SAME LEAF NODE GLOTTOCODE THAT
HAVE DIFFERENT PHONEMES / NUMBERS OF PHONEMES.

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(knitr)
library(ggplot2)

url_ <- "https://github.com/phoible/dev/blob/master/data/phoible.csv?raw=true"
phoible <- read.csv(url(url_), stringsAsFactors=FALSE)
```

Below is a summary of the number of isocodes that are represented by multiple
inventories in PHOIBLE:

```{r stemplot_inv_per_iso}
phoible %>%
    group_by(ISO6393) %>%
    summarise(y=n_distinct(InventoryID)) %>%
    count(y) %>%
    ggplot(aes(y=as.factor(y), x=n)) +
    geom_point() +
    geom_segment(aes(x=0, xend=n, y=as.factor(y), yend=as.factor(y))) +
    scale_x_log10() +
    theme_light() +
    labs(title="Prevalence of multiple inventories per ISO code",
         x="Number of ISO codes having N inventories (log scale)",
         y="N inventories")
```

So most ISO codes have only 1 inventory (the long, bottom line). What is the ISO
code that has 37 inventories?!

```{r}
phoible %>%
    group_by(ISO6393) %>%
    summarise(n_inventories=n_distinct(InventoryID)) %>%
    arrange(desc(n_inventories)) %>%
    filter(n_inventories > 5) %>%
    kable()
```

Turns out there are 37 inventories with `NA` as their isocode.
XXX TODO ADD A SENTENCE ABOUT WHY THEY ARE `NA` XXX.

Here is the same representation, for glottocodes instead of isocodes:

```{r stemplot_inv_per_glotto}
phoible %>%
    group_by(Glottocode) %>%
    summarise(y=n_distinct(InventoryID)) %>%
    count(y) %>%
    ggplot(aes(y=as.factor(y), x=n)) +
    geom_point() +
    geom_segment(aes(x=0, xend=n, y=as.factor(y), yend=as.factor(y))) +
    scale_x_log10() +
    theme_light() +
    labs(title="Prevalence of multiple inventories per glottocode",
         x="Number of glottocodes having N inventories (log scale)",
         y="N inventories")
```

Again, most glottocodes are represented by just one or two inventories in
PHOIBLE. Let's see what those few glottocodes are that have so many inventories:

```{r}
phoible %>%
    group_by(Glottocode) %>%
    summarise(n_inventories=n_distinct(InventoryID)) %>%
    arrange(desc(n_inventories)) %>%
    filter(n_inventories > 5) %>%
    kable()
```


# Filtering and sampling inventories

If multiple inventories per isocode/glottocode are problematic for your
analysis or research question, one approach is to select one inventory from each
isocode/glottocode via random sampling:

```{r}
phoible %>%
    group_by(Glottocode) %>%
    sample_n(1) ->
    one_inv_per_glottocode

phoible %>%
    group_by(ISO6393) %>%
    sample_n(1) ->
    one_inv_per_isocode

message("Picking one inventory per glottocode reduces PHOIBLE from ",
        n_distinct(phoible$InventoryID), " inventories\nto ",
        nrow(one_inv_per_glottocode),
        " inventories. Picking one per ISO 639-3 code yields ",
        nrow(one_inv_per_isocode), " inventories.")
```

Another approach is to only use only inventories from a data source that already
provides a one-inventory-per-language sample. For example, UPSID represents a
"quota" sample (one language per family, for some definition of "family"). To do
this, you can filter the PHOIBLE data by the `Source` column:

```{r}
phoible %>%
    filter(Source == "upsid") ->
    upsid

# show that there is exactly one inventory per ISO 639-3 code:
upsid %>%
    group_by(ISO6393) %>%
    summarise(InventoryID=n_distinct(InventoryID)) %>%
    select(InventoryID) %>%
    distinct() %>%
    kable()
```

However, note that the state of knowledge of linguistic genealogical groupings
has changed since UPSID was published. A quota sample that reflects current
understanding can also be achieved. Here, we choose one leaf node from within
each top-level Glottolog family node that has representation in PHOIBLE:

```{r}
# TODO XXX EXAMPLE OF QUOTA SAMPLE USING GLOTTOLOG XXX
# caveat: every Isolate will be included
```

Another approach is to select inventories based on properties of the inventories
themselves, such as whether they include
representations of allophones or contrastive tone, etc.

```{r}
# TODO XXX EXAMPLE OF PREFERENTIAL SOURCE ORDERING HERE XXX
```

If you're inclined to be a "splitter", you might prefer to pick the largest
inventory for a given isocode...

```{r}
# TODO XXX EXAMPLE OF PICKING AMONG DUPLICATES BY INVENTORY SIZE XXX
```

